Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-17 paper-1/manuscript.tex
Latex fixes & minor edits

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{}
    \label{fig1}
\end{figure*}
\textbf{Fig. 1} Model and paradigm. \textbf{a} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes, illustrated above the plots. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Inputs to the (post-)synaptic conductances $g_e$ and $g_i$ (vertical bars), and the STD depression variable $x_{Exc}$ of the excitatory presynaptic neuron (dashed line). Note that excitatory inputs are equal to $U x_{Exc} w$ as per Equation \ref{eqn-gsyn}. Middle: Excitatory and inhibitory synaptic conductances ($g_e$ and $g_i$, respectively). Bottom: Membrane voltage, firing threshold baseline (gray) and adaptive threshold, demonstrating TA in the excitatory postsynaptic neuron. \textbf{b} Sample network layout, showing the spatial location of excitatory and inhibitory neurons, and all outgoing synaptic connections from two neurons of each type. \textbf{c} Distance from stimulation site 1 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (large circles), which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{d} Illustration of the stimulation paradigm. Each marker represents a stimulation, with the stimulus identity indicated by color. The target stimulus (A) is represented as solid red, and non-target stimuli are represented with faded colors. Three randomized sequences were presented, containing 80\% A and 20\% B (\std{}), 20\% A and 80\% B (\dev{}), and 20\% of each of 5 stimuli, including A and B (\msc{}). Note that, while the onset asynchrony of 500 ms is represented faithfully, the horizontal extent of the stimulation markers has no meaning; stimuli were presented instantaneously rather than over an extended period.

@@

Throughout this paper, we refer to three main quantities of interest: the stimulus response $\R{}{}$ defined below, the threshold adaptation value $\T{}{}$ defined in Equation \ref{eqn-TA}, and the short-term depression effect $\D{}{}$, which is introduced later in Equation \ref{eqn-D}. We employ sub- and superscripts to identify the precise quantity as follows, using $\R{}{}$ for illustration: $\R[seq]{X}{p}$ refers to the response of neuron(s) $p$, averaged across all trials of type $X$ in the sequence $seq$. Where $p$ does not identify individual neurons (e.g., $i$ or $j$), but rather a subpopulation (e.g., $exc$) or the whole network (no index), this refers to the median (of the individual quantity $\R[seq]{X}{i}$) across the indicated set of neurons. The trial type $X$ is either $A$ or $B$, referring to trials with stimulation at these sites, or $N$, indicating all \textbf{n}on-target trials B through E. Finally, we use the notation $\Delta \R{}{}$ to refer to the sequence contrast $\R[dev]{}{} - \R[msc]{}{}$, and $\mean{\R{}{p}}$ to indicate the mean, rather than median, of the individual quantities in a population $p$.

We define the response $\R{X}{i}$ as the mean spike count of neuron $i$ across time bins in trials of type $X$. Almost equivalently, we define $\T{X}{i}$ (and $\D{}{}$ analogously) as the mean across trials of type $X$ of $\T{}{i}(t=0)$, i.e., the average TA voltage at the beginning of trials, immediately before stimulation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model slightly enhances the effects of TA.
\textbf{a} Reductions in mean response size $\mean{\R{A}}$ in each sequence as a consequence of adding STD to the model, across networks and stimuli.
\textbf{b} $\Delta \R{N}{}$ (top) and $\Delta \T{B}{}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{c} $\Delta \R{N}{exc}$ across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$ across excitatory neurons. Compare to Figure \ref{fig5}d.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that $\T{A}$ tends to be reduced more in the full model.
\textbf{f} $\Delta \T{A}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ across excitatory neurons. Compare to Figure \ref{fig4}e.
\textbf{g} $\Delta \R{A}{}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

@@

We note that the direct and indirect effects of STD on global suppression appear to be on the same order of magnitude. To see this, compare the change in $\Delta \T{A}{}$ as a result of adding STD to the model (Figure \ref{fig7}e, bottom) with the corresponding $\Delta \D{A}$ (Figure \ref{fig8}e), both of which are on the order of $\approx$ 0.1 mV. Conversely, notice that, since TA and STD work in similar fashion to increase the \dev{} target response, it is very plausible that TA likewise increases the effect of STD, by enhancing the STD-mediated local suppression of B and thereby increasing the contrast in STD-mediated global suppression. Thus, this synergy allows the individually weak STD to exert a significantly larger effect when paired with the stronger TA.

@@

\begin{figure}%
    \centering
    \includegraphics{sfig-STD-correlation-decay.pdf}
    \caption{}
    \label{sfig-STD-correlation-decay}
\end{figure*}
\textbf{Figure S?} The STD variable $x_i$ only correlates with the activity of the immediately preceding trial. Each marker represents the coefficient $\rho$ of the Pearson correlation between the spike counts in a target trial and the corresponding values $x_i$ at the beginning of the immediately following trial ($\Delta$ trials = 1) or after one intervening trial ($\Delta$ trials = 2). Recall that $x$ is reduced upon spiking, which implies that an effective correlation is negative. Color represents the p-value of the alternative hypothesis that $\rho < 0$. Horizontal lines indicate the median $\rho$ and p. Horizontal jitter is introduced to aid visual separability of the markers.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{}
    \label{fig1}
\end{figure*}
\textbf{Fig. 1} Model and paradigm. \textbf{a} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes, illustrated above the plots. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Inputs to the (post-)synaptic conductances $g_e$ and $g_i$ (vertical bars), and the STD depression variable $x_{Exc}$ of the excitatory presynaptic neuron (dashed line). Note that excitatory inputs are equal to $U x_{Exc} w$ as per Equation \ref{eqn-gsyn}. Middle: Excitatory and inhibitory synaptic conductances ($g_e$ and $g_i$, respectively). Bottom: Membrane voltage, firing threshold baseline (gray) and adaptive threshold, demonstrating TA in the excitatory postsynaptic neuron. \textbf{b} Sample network layout, showing the spatial location of excitatory and inhibitory neurons, and all outgoing synaptic connections from two neurons of each type. \textbf{c} Distance from stimulation site 1 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (large circles), which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{d} Illustration of the stimulation paradigm. Each marker represents a stimulation, with the stimulus identity indicated by color. The target stimulus (A) is represented as solid red, and non-target stimuli are represented with faded colors. Three randomized sequences were presented, containing 80\% A and 20\% B (\std{}), 20\% A and 80\% B (\dev{}), and 20\% of each of 5 stimuli, including A and B (\msc{}). Note that, while the onset asynchrony of 500 ms is represented faithfully, the horizontal extent of the stimulation markers has no meaning; stimuli were presented instantaneously rather than over an extended period

@@

Throughout this paper, we refer to three main quantities of interest: the stimulus response $\R{}{}$ defined in the next paragraph, the threshold adaptation value $\T{}{}$ defined in Equation \ref{eqn-TA}, and the short-term depression effect $\D{}{}$, which will be introduced later in Equation \ref{eqn-D}. We employ sub- and superscripts to identify the precise quantity as follows, using $\R{}{}$ for illustration: $\R[seq]{X}{p}$ refers to the response of neuron(s) $p$, averaged across all trials of type $X$ in the sequence $seq$. Where $p$ does not identify individual neurons (e.g., $i$ or $j$), but rather a subpopulation (e.g., $exc$) or the whole network (no index), this refers to the median (of the individual quantity $\R[seq]{X}{i}$) across the indicated set of neurons. The trial type $X$ is either $A$ or $B$, referring to trials with stimulation at these sites, or $N$, indicating all \textbf{n}on-target trials B through E. Finally, we use the notation $\Delta \R{}{}$ to refer to the sequence contrast $\R[dev]{}{} - \R[msc]{}{}$, and $\mean{\R{}{p}}$ to indicate the mean, rather than median, of the individual quantities in a population $p$.

We define the response $\R{X}{i}$ as the mean spike count of neuron $i$ across time bins in trials of type $X$. Almost equivalently, we define $\T{X}{i}$ (and $\D{X}{i}$ analogously) as the mean across trials of type $X$ of $\T{}{i}(t=0)$, i.e., the average TA voltage at the beginning of trials, immediately before stimulation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model slightly enhances the effects of TA.
\textbf{a} Reductions in mean response size $\mean{\R{A}{}}$ in each sequence as a consequence of adding STD to the model, across networks and stimuli.
\textbf{b} $\Delta \R{N}{}$ (top) and $\Delta \T{B}{}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{c} $\Delta \R{N}{exc}$ across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$ across excitatory neurons. Compare to Figure \ref{fig5}d.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that $\T{A}{}$ tends to be reduced more in the full model.
\textbf{f} $\Delta \T{A}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ across excitatory neurons. Compare to Figure \ref{fig4}e.
\textbf{g} $\Delta \R{A}{}$ across networks and stimuli in the full and ablated models, arranged as in panel c. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

@@

We note that the direct and indirect effects of STD on global suppression appear to be on the same order of magnitude. To see this, compare the change in $\Delta \T{A}{}$ as a result of adding STD to the model (Figure \ref{fig7}e, bottom) with the corresponding $\Delta \D{A}{}$ (Figure \ref{fig8}e), both of which are on the order of $\approx$ 0.1 mV. Conversely, notice that, since TA and STD work in similar fashion to increase the \dev{} target response, it is very plausible that TA likewise increases the effect of STD, by enhancing the STD-mediated local suppression of B and thereby increasing the contrast in STD-mediated global suppression. Thus, this synergy allows the individually weak STD to exert a significantly larger effect when paired with the stronger TA.

@@

\begin{figure*}%
    \centering
    \includegraphics{sfig-STD-correlation-decay.pdf}
    \caption{}
    \label{sfig-STD-correlation-decay}
\end{figure*}
\textbf{Figure S?} The STD variable $x_i$ only correlates with the activity of the immediately preceding trial. Each marker represents the coefficient $\rho$ of the Pearson correlation between the spike counts in a target trial and the corresponding values $x_i$ at the beginning of the immediately following trial ($\Delta$ trials = 1) or after one intervening trial ($\Delta$ trials = 2). Recall that $x$ is reduced upon spiking, which implies that an effective correlation is negative. Color represents the p-value of the alternative hypothesis that $\rho < 0$. Horizontal lines indicate the median $\rho$ and p. Horizontal jitter is introduced to aid visual separability of the markers.
